---
title: "Parsing ABC Tunes with `rly`"
author: "Curtis C. Bohlen"
date: "11/14/2021"
output:
  github_document:
    toc: true
    fig_width: 5
    fig_height: 4
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(fig.align = 'center',
                      fig.width = 5, fig.height = 4,
                      collapse = TRUE, comment = "#>")
```

# Introduction
In this notebook, I explore the R parsing engine `rly`.  `rly` is a port to R of 
a python package, `ply`. The name `ply` implies "python lex and yacc", where
"lex" and "yacc" are classic UNIX tools for building formal parsers. 

Here we develop partial parsers and lexers for ABC Tunes, to develop ideas 
and constraints.

# A Slightly More Complex Example

This lexer and parser accept a string of notes, with white space and other
complexities, so it is one step more complex than the one we just developed.

# Test Data
```{r}
arpeggios_1 <- "CEG2 CFA2 DGB2"

arpeggios_2 <- "GBD2 GCE2 AD^F2"   # Add a Sharp

arpeggios_3 <- "G_BD2 GC_E2 ADF2"   # Add two flats

arpeggios_4 <- "GBd2 Gce2 Ad^f2"   # Add variation in octaves

arpeggios_5 <- "GBD'2 GC'E'2 AD'^F'2"   # octave modifiers should produce same as last

arpeggios_6 <- "g,b,d2 g,ce2 a,d^f2"   # octave modifiers should produce same as last

arpeggios_7 <- "CEGz CFAz DGBx x4 Z2 " # Testing rests.
```


# A Simple Tune Lexer
```{r}
TOKENS = c('ACCIDENTALS', 'BASENOTE', 'OCTAVE_ADJ', 'DURATION', 'WHITESPACE')
LITERALS = c()

TestLexer <- R6::R6Class("Lexer",
  public = list(
    tokens = TOKENS,
    literals = LITERALS,
  
    # The ABC standard allows multiple accidental markers, possibly mixed
    t_ACCIDENTALS = '[\\^=_]+',
    t_BASENOTE = "[a-gA-G]",
    t_OCTAVE_ADJ = "[\\',]+",
    
    t_DURATION = function(re='[\\/]?\\d+', t) {
         first <- substr(t$value, 1, 1)
         if (first == '/') {
           rest <- sub('.', '', t$value)  # strip first character
           t$value <- 1 / strtoi(rest)    # Convert to integer and invert
         } else {
           t$value <- strtoi(t$value)
         }
         return(t)
    },

    t_WHITESPACE = "\\s+",
    
    # we Need a Function to Increment Line Numbers, to help withe fixing errors
    t_newline = function(re='\\n+|(?:\\n\\r)+', t) {         
      t$lexer$lineno <- t$lexer$lineno + nchar(t$value)
      return(NULL)
    },
    
    # define an error function for unrecognized tokens or characters,
    # here we simply skip to teh next character and try again.
    t_error = function(t) {
      cat(sprintf("Illegal character '%s'", t$value[1]))
      t$lexer$skip(1)
      return(t)
    
    }
  )
)

testlexer <- rly::lex(TestLexer)
```


We use a parser by feeding it some input (with the `$input()` method) then 
pulling tokens out one at a time(with the `$token()` method).
```{r}
testlexer$input(arpeggios_1)
while (!is.null(current_token <- testlexer$token())) {
  print(current_token)
}
```



# A Simple Tune Parser
Next, we build a simple parser to take the output from a lexer and turn in into
something more useful.  This parser builds on one developed in another 
notebook.  The changes add code for tracking a few additional features of
real music, including rests, slurs and ties.

The other major change is use of an empty production to simplify coding 
of optional components of hte standard.

```{r}
TestParser <- R6::R6Class("Parser",
  public = list(
    tokens = TOKENS,  # defined above!
    literals = LITERALS,
    # Parsing rules
    
    p_note_seq = function(doc = 'note_seq : note note_seq
                                          | spaced_note note_seq', p) {
        p$set(1, append(p$get(3), list(p$get(2)), after = 0))  # probably slow....
    },
    
    p_note_seq_start = function(doc = 'note_seq : note
                                                | spaced_note', p) {
        p$set(1, list(p$get(2)))
    },
    
    # the following could be folded back int onote, but eventually 
    # we will want to handle this as a special case too...
    p_spaced_note = function(doc = 'spaced_note : note WHITESPACE', p) {
      p$set(1, p$get(2))
    },
    
    p_note  = function(doc = 'note : note_dur
                                   | note_oct
                                   | note_acc
                                   | basenote', p) {
        p$set(1, p$get(2))
    },
    
    # If the note has a duration specified, we need to add that now.
    # we could be working with three possible precursors
     p_note_duration  = function(doc = 'note_dur : note_oct DURATION
                                                 | note_acc DURATION
                                                 | basenote DURATION', p) {
        note <- p$get(2)
        note$duration <- p$get(3)
        p$set(1, note)
    },
    
    
    # If we have octave adjustments, they are signaled after the basenote
    # but now we need to recognize that we could be working with a bare
    # basenote, or one modified by accidentals
    p_note_oct = function(doc = 'note_oct : note_acc OCTAVE_ADJ
                                          | basenote OCTAVE_ADJ', p) {
        note <- p$get(2)
        chars <- p$get(3)
        chars <- strsplit(chars, '')[[1]]  # strsplit is vectorized
        octave <- note$octave
        for (n in seq_along(chars)) {
          if (chars[n] == "'") {
            octave = octave + 1
          } else if (chars[n] == ',') {
            octave = octave - 1
          }
        }
        note$octave <- octave
        p$set(1, note)
    },
    
    
    # If we have one or more accidentals, they come before the base note
    # the ABC standard allows multiple accidentals
    p_note_acc=  function(doc = 'note_acc : ACCIDENTALS basenote', p) {
      note <- p$get(3)
      chars <- p$get(2)
      chars <- strsplit(chars, '')[[1]]  # strsplit is vectorized
      accidental <- note$accidentals

      for (n in seq_along(chars)) {
        if (chars[n] == '^') {
          accidental = accidental + 1
        } else if (chars[n] == '_') {
          accidental = accidental - 1
        }
      }
      note$accidentals <- accidental
      p$set(1, note)
    },
    
    # we start with basenote, to create a new note data structure, which we modify
    # as we add other components to determine true_pitch and pitch duration.
    p_basenote    = function(doc = 'basenote : BASENOTE', p) {
      char <- p$get(2)
      if (char == toupper(char)) {
        octave <- 4
      } else {
        octave <- 5
      }
      # create our note data structure
      p$set(1, new_note(name = toupper(char), accidental = 0, 
                        octave = octave, duration = 1))
    },
    
    p_error = function(p) {
      if(is.null(p)) cat("Syntax error at EOF")
      else           cat(sprintf("Syntax error at '%s'", p$value))
    }
  )
)
testparser <- yacc(TestParser, debug = TRUE)

```

# Testing the Parser
```{r}
p <- testparser$parse(arpeggios_1, testlexer)
p

```

It's not too hard to assemble that into a dataframe.  The only trick is that 
we have to unclass the note objects. (Alternatively, we could define an 
`as.data.frame.note()` function that does that for us).

```{r}
do.call(rbind, lapply(p, function(note) data.frame(unclass(note))))
```

Or  Flatten the list of notes to a vector representation of the strings, and show it.
```{r}
v <- vector(length = length(p))
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```

```{r}
p <- smallparser$parse(arpeggios_2, smalllexer)
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```

```{r}
p <- smallparser$parse(arpeggios_3, smalllexer)
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```

```{r}
p <- smallparser$parse(arpeggios_4, smalllexer)
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```

```{r}
p <- smallparser$parse(arpeggios_5, smalllexer)
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```

```{r}
p <- smallparser$parse(arpeggios_6, smalllexer)
for (n in seq_along(p)) {
  cat(format(p[[n]]))
  cat('\n')
}
```


