---
title: "Exploring Music Packages with R"
output: html_notebook
---

# Package `music`
https://cran.r-project.org/web/packages/music/index.html

This package includes functions for building note, chords, progressions, and 
scales.  It manages sound by sending wav files to the `audio` package.

```{r}
library(tidyverse)
library(music)
```


I found no related vignettes, so we need to work through specific examples.

I start with `playNote()`, since it is likely to be something we want to use 
regularly.

The duration settings do not appear to work if we set different durations for 
each note.  While the function signature implies that we could offer multiple
values here, I fond that that only works if they are all the same, in which case
we can just as easily use a single 
```{r}
playNote(c('A4', 'B4', 'C5'), duration = 2)
```


Note that the data structure implied here splits pitches and durations.

Pitches are described with conventional pitch names, with octaves defined as 
usual, with C4 = middle c, and A4 = 440 (although that can be changed with a 
function).  Pitch names are case sensitive,

```{r}
playNote(c('A4', 'Bb4', 'B4', 'C5', 'C#5'), duration = 2)
```

WE can play a quasi-tune, but only if all notes are the same duration.
```{r}
playNote(c('B4','A4','G4','A4','B4','B4','B4'), duration = 1)
```

## Building a tune
Since we can't play a tune directly, we need to  use a loop of 
some sort, but this does not work terribly well. Sounds run together without 
`Sys.sleep()`.... But with it you hear clicks and pops, and rythmn is messed up.

```{r}
mary <- tibble(notes = c('B4','A4','G4','A4','B4','B4','B4','A4','A4','A4','B4','D5','D5'), 
               durations = c(1,1,1,1,1,1,2,1,1,2,1,1,2))

for(row in seq(1,length(mary$notes))) {
  playNote(mary$notes[[row]], duration = mary$durations[[row]])

  Sys.sleep(0.75)
}

```

# Package `audio`
```{r}
library(audio)
```
Audio is a package that plays sounds.
```{r}
play(sin(1:10000/20))
```

But the documentation is very sparse.  One could potentially assemble a tune 
directly from sequential sound samples or wave forms, but that would be labor
intensive.  This is apparently the package that `music` uses to play notes
and chords.

# Package `tuneR`
```{r}
library(tuneR)
```

Package `tuneR` contains functions to read several music formats(wav mp3 and
midi, at least) and analyze and manipulate WAV files.  It contains many 
utilities for manipulating, analyzing, resampling, or displaying WAV files.

It appears to be a favorite for people working with complex sound in R, 
including birdsong and ambient sounds.

It also allows wav tunes to be converted to text files that can be presented as
sheet music using music typesetting software LilyPond (open source / GNU
package).

This introduces yet another music notation system.... The function expects
notes to be passed as a data frame with four components. Note, duration, 
punctate, and slur. But the function is considered under development.

```{r}
noteFromFF(600, diapason = 440, roundshift = 0)
```
So, that gives us a note 5 halftones above A
Formula is this: round(12 * log(x / diapason,2) + roundshift

```{r}
notenames(c(12, 0,  -12, -24))
notenames(c(0, 4, 7))
```

Note that the note annotation looks suspiciously similar to ABC, in terms of
representing octaves, but not accidentals or duration.

I found a nice examples of using the package for music theory:
https://www.r-bloggers.com/2011/10/the-psychology-of-music-and-the-%E2%80%98tuner%E2%80%99-package/

It's a bit out of date, but minor tweaks are interesting.

```{r}
sound <- sine(440, bit = 32)
 
writeWave(sound, '440.wav')
```

```{r}
sound <- normalize(bind(sine(440, bit = 32),
              sine(880, bit = 32),
              sine(440, bit = 32) + sine(880, bit = 32)))

writeWave(sound, 'octaves.wav')
```

Note that the author of that blog mantains a github archive for generating
the frequencies we need to listen to different types of temperament / tunings.

# Package `gm`
```{r}
library(gm)
```

Package `gm` is a package for "generating music" according to a "grammar of
music". It appears to accomplish several things via a connection with
MuseScore.  For `show()` in particular, `gm` appears to translate the music
tMusicXML, then pass that to MuseScore to be rendered.

On the whole, the internal representations used are solid, and the approach to 
"adding up" music parts may suit our needs.

```{r}
m <- Music()
m <- m +
  # add a 4/4 time signature
  Meter(4, 4) +
  # add a musical line of a C5 whole note
  Line(pitches = list("C5"), durations = list("whole"))

m

```
The program's ability to generate scores is VERY slow, and scores do not
appear in the markdown document, but in the plot Viewer.  The following takes 
perhaps 30 seconds to render a single note.
```{r}
show(m, to = c('score'))
```

Generating audio is even slower.  Again, the output turns up in the viewer, 
whence the user can play it. Again, you'll wait close to a minute for output 
to appear.
```{r}
show(m, to = c('audio'))
```
Note that we can add notes.  I am uncertain what makes the decision to add these 
as bass clef.
```{r}
m <- m + Line(
  pitches = list("C3", "G3"),
  durations = list("half", "half")
)

show(m)
```

Adding notes as a voice adds them to  a specified line. Note the use 
of `c()` to specify multiple notes sounding simultaneously.

```{r}
m <- m + Line(
  pitches = list(c("E4", "G4"), c("D4", "F4"), c("F4", "A4"), c("E4", "G4")),
  durations = list("quarter", "quarter", "quarter", "quarter"),
  as = "voice",
  to = 1)

show(m)
```

The underlying data structure is a list of lists, with various nested structures.
```{r}
str(m)
```

The similarity to MusicXML is evident, but not complete.

`gm` uses either pitch names or MIDI pitch numbers.  Names are defined as
follows:

1.  a tone name (from A to G),  
2.  an optional accidental (-, --, # or ##), and  
3.  a number identifying the pitch’s octave (from 0 to 9).  

Durations are principally given by "duration types" like "whole", 'half', and 
'32nd'. Most words can be abbreviated to first letters.  You can add zero to 
four dots....  There is special tuplet notation.

The package points to a conversion table between MIDI pitch numbers and names
defined in this way.  That is, interestingly, in a Wikipedia  article about 
'Scientific Pitch Notation".

That includes the following useful formulae:
Mathematically, given the number n of semitones above middle C, the fundamental
frequency in hertz is given by 440 ⋅ 2^ ( n − 9 ) / 12.  (see twelfth root of two)

$$\displaystyle 440\cdot 2^{(n-9)/12}$$


Given the MIDI Note number m, the frequency of the note is normally 
440 ⋅ 2 ( m − 69 ) / 12
$$\displaystyle 440\cdot 2^{(m-69)/12}$$
Hz, using standard tuning. 

# Package `tabr`
Includes music analysis functions.  The main purpose is "to manipulate musical 
information in notation form." Includes a music syntax converters from one
music data representation to another.  Including tools for making tabs and
export sheet music with LilyPond. Emphasis is on guitar tabs.

```{r}
library(tabr)
```


A central component is the "noteworthy" string -- a text format for music data.
These strings have some variations on how they are shown.

```{r}
x <- "a# b_*2 c, d'' e3*2 g_4 c2e_2g2*2"
x <- as_noteworthy(x)
x
```
```{r}
summary(x)
```
```{r}
notes <- "c d e f g a b ceg~ ceg"
info <- "8*8 1"
x <- as_music(notes, info)
x
```
```{r}
summary(x)
```
```{r}
music_split(x)
```

```{r}
x <- "a,4*5 b,4- c4 cgc'e'~4 cgc'e'1 e'4 c' g c ce'1"
musical(x)
x <- as_music(x)
x
```

```{r}
str(x)
```




#Package `chorrrds`
Extracts chords from the Cifraclub website to download and organize music chords.  Cifraclub is a spanish language website that contains information of chords and lyrics for popular music.

The packages primary focus is a handful of functions for analysis of chord progressions.  It includes things like
chord distance, and chord ngrams.

The underlying data structure is a fairly simple text depiction of chords using standard guitar chord conventions.
For hte most part, however, the package does not interpret the strings as musical sequences of notes.  It uses '#' for sharps and 'b' for flats.
```{r}
library(chorrrds)
```


```{r}
q <- get_songs('Queen')

chords <- get_chords(q$url[9])
chords
```

```{r}
ng <- chords_ngram(chords, n = 3)
ng
```

```{r}
ng %>%
  arrange(chords_ngram)
```


# Package `music21`
A R wrapper around python's `music21` package.  This will require some 
finagling to get it to work on correctly on windows.
The package does not appear on CRAN, so you need to load it from GitHub.

Like some other packages, the plot and view methods depend on the open source 
package LilyPond. Ironically, LilyPond wants its own Python...

Running music21 is less than trivial, since the package needs to find both 
Python and the Python music21 module.

## Setting Up `reticulate`

I set up a conda environment as follows:

```
>conda create -n music21
>conda activate music21
>conda install pip scipy pandas matplotlib spyder
```

The third line also installs python, and many commonly used python modules.

`music21` includes several other dependencies that are not yet satisfied.  Conda documentation suggests installing as many dependencies as possible using `conda`, then using `pip` only to install the very last (un conda) pieces. The following unsatisfied dependencies were added when I first tried to install `music21` into a minimal conda environment.

```
>conda install more-itertools jsonpickle joblib chardet 
```

That results in just a single unsatisfied dependency other than `music21` 
installed via `pip`, `webcolors`, which was not available through my standard 
conda channels.

```
>pip install music21 webcolors
```

## Project-Specific Python
After some frustration trying to check and set the 'RETICUALTE_PYTHON' 
environment variable, I figured out that although a direct call to 
`Sys.getenv()` did not show a 'RETICUALTE_PYTHON' variable set, it was somehow
getting set in the background.  By default, it was being set to the default
Anaconda installation, where the `music21` module is NOT installed, so
everything failed.

If 'RETICUALTE_PYTHON' is set, you can not override the selection of python
versions or environments using `reticulate::use_condaenv()` or other tools.

I changed the project Python by specifying it in the "Project Options" 
dialog, under Python, in RStudio, and selecting the `correct`music21` conda 
environment, and everything works.

If you are not using RStudio in Windows, your mileage may vary.

```{r}
library(reticulate)
```

```{r}
Sys.getenv('RETICUALTE_PYTHON')
```

```{r}
py_discover_config(use_environment = 'music21')
```

Note the final note, telling me that DESPITE not finding a 'RETICUALTE_PYTHON'
environment variable,  the specific Python WAS determined by an environment 
variable.....  Sigh....

We still want to tell `reticulate` where to look for relevant python packages.

```{r}
use_condaenv("music21", required = TRUE)

library(music21)
music21
```

The music21 object stores the python module. It is possible to run any function from it just by using $, treating it as a Reference Class object.

```{r}
note <- music21$note$Note("C#")
note
```

To plot the note, we also need to tell `music21` where to find the LilyPond
executable.  I installed this on windows. This is slightly tricky, since it is 
set back in python land, not R land.
THe executable is located at:

C:\Program Files (x86)\LilyPond\usr\bin\lilypond.exe

```{r}
m21 <- import("music21")
us <- m21$environment$UserSettings()
us
```

Note that several of these paths are wonky, being redrawn as relative to the
home directory of this project.

so far, so good, but I can't actually change those values in R.  What appears 
to be happening is that we are changing the values in R memory, but when we look 
at the whole structure  (which mirrors a configuration file) the changes are not
recorded.

```{r}
us$lilypondPath <- ""
us$lilypondPath
us
```

So it looks like we have to do the work in Python land. WE then check if it
worked back in R land.

```{r}
py_run_string("us2 = m21.environment.UserSettings()")
py_run_string("us2['lilypondPath'] = 'C:/Program Files (x86)/LilyPond/usr/bin/lilypond.exe'")


us <- m21$environment$UserSettings()
us
```

```{r}
py_run_string("print(us2['lilypondPath'])")
```

```{r}
note %>% 
  plot()
```


# r-music on github
https://github.com/r-music
provides links to several R music-related packages.  Note clear what the deal is....


# MusicXML


