---
title: "Reticulate and `music21`"
output: html_notebook
---

# Package `music21`
A R wrapper around python's `music21` package.  Likely to require some 
finagling to get it to work on correctly on windows.
The package does not appear on CRAN, so you need to load it from GitHub.

Like some other packages, the plot and view methods depend on the open source 
package LilyPond.

Running music21 is less than trivial, since the package needs to find both 
Python and the Python music21 module.


# Setting Up a Conda Envrionment
I set up a conda environment as follows:

```
>conda create -n music21
>conda activate music21
>conda install pip scipy pandas matplotlib spyder
```

The third line also installs python, and many commonly used python modules. Here
it is a version of python 3.9, more recent than the 3.7 in my base conda 
environment.

`music21` includes several other dependencies that are not yet satisfied.  Conda
documentation suggests installing as many dependencies as possible using
`conda`, then using `pip` only to install the very last (un conda) pieces. The
following unsatisfied dependencies were added when I first tried to install
`music21` into a minimal conda environment. (These may be outdates, since I 
added spyder to be able to work directly with music21.)

```
>conda install more-itertools jsonpickle joblib chardet 
>pip install music21
```
That results in just a single unsatisfied dependency other than `music21` 
installed via `pip`, `webcolors`, which was not available through my standard 
conda channels.

# Project-Specific Python
It appears RStudio is setting an envrionment variable behind the 
scenes.  After some frustration trying to check and set the 'RETICUALTE_PYTHON' 
environment variable, I figures=d out that although a direct call to 
`Sys.getenv()` did not show a 'RETICUALTE_PYTHON' variable set, it was somehow
getting set in the background.  By default, it was being set to the default
Anaconda installation, where I had not installed the `music21` module, so
everything failed.

The problem is, if 'RETICUALTE_PYTHON' is set, I can not override the selection
of envrionments using `reticulate::use_condaenv()` or other tools.

I changed the project Python by specifying it in the "Project Options" 
dialog, under Python, and selecting the correct conda environment.

# Checking the 'RETICUALTE_PYTHON' variable
```{r}
Sys.getenv('RETICUALTE_PYTHON')
```

# Testing `reticulate`
```{r}
library(reticulate)
```

```{r}
Sys.getenv('RETICUALTE_PYTHON')
```

If we don't specify anything, we get the python 3.7 in our base Anaconda
installation, but since we set the Project Python, python will find the 
version in the requested conda envrionment
```{r}
py_discover_config()
```
Note the final note, telling me that DESPITE not finding a 'RETICUALTE_PYTHON'
environment variable,  the specific Python WAS determined by an environment 
variable.....  Sigh....

# Running Music21
We still want to tell `reticulate` where to look for relevant python packages.

```{r}
use_condaenv("music21", required = TRUE)
```

```{r}
library(music21)
music21
```

The music21 object stores the python module. It is possible to run any function from 
it just by using $, treating it as a Reference Class object.

```{r}
note <- music21$note$Note("C#")
note
```


But we still have trouble finding lilypond.

Note that some online references suggest the problem may be that python or 
music21 has trouble with paths with spaces, so we uninstalled and reinstalled 
it at 
`C:\LilyPond\usr\bin\lilypond.exe`


```{r error = TRUE}
plot(note)
```

I have not been able to solve this - -althoug hI have lilyPond loaded, R can't 
find it.


## Changing Environment Variables
The lilypondPath is often set inappropriately.  Once set in the env variables, 
however, it is persistent between
```{r}
us <- music21$environment$UserSettings()
us[['lilypondPath']]
```

But this proves difficult to change in R
```{r}
#us[['lilypondPath']] <- 'C:\\LilyPond\\usr\\bin\\lilypond.exe'
#us[['lilypondPath']]
```
 
As far as I can tell, the R version of the data is a copy of the Python data
structure, so we need to alter the Python data structure, as follows:
```{r}
py_run_string("import music21")
py_run_string("import sys")
py_run_string("modulenames = set(sys.modules) & set(globals())")
py$modulenames
```

```{r}
py_run_string("us2 = music21.environment.UserSettings()")
py_run_string("us2['lilypondPath'] = r'C:\\LilyPond\\usr\\bin\\lilypond.exe'")
py_run_string("us2['musicxmlPath'] = r'C:\\Program Files\\MuseScore 3\\bin\\MuseScore3.exe'")
py_run_string("us2['musescoreDirectPNGPath'] = r'C:\\Program Files\\MuseScore 3\\bin\\MuseScore3.exe'")

```

```{r}
us <- music21$environment$UserSettings()
us[['lilypondPath']]
us[['musicxmlPath']]
```


```{r error = TRUE}
plot(note)
```

# At least MuseScore works as a back end!
An alternate form, bote.show(), pulls in MuseScore to display the note, so
perhaps the lack of LilyPond in not so great.  Starting up MuseScore, however,
is fairly slow.
```{r}
note$show()
```

Bun honestly, why not just work with Music21 directly in Python?  That is 
probably a much more reasonable approach.

Note that under the hood, music21 likes MusicXML format music ata a fai amount.

