---
title: "Load Example ABC data to Text"
output: html_notebook
---


## code to prepare `DATASET` dataset goes here

```{r}
library(tidyverse)
library(rlang)
```

We assemble a list of file names with  ".abc" in them.  Presumably these are
ABC music files, with the ".abc' file extension (although we don't test that 
explicitly).
```{r}
parent <- dirname(getwd())
foldernm<-  'inst/extdata'
folder <- file.path(parent, foldernm)
fns <-list.files(folder, ".abc")

examples <- list()
for (fn in fns) {
    nm <- substr(fn, 1, nchar(fn)-4)
    lines <- readLines(file.path(folder, fn), skipNul = TRUE)
    if(! grepl('^%abc', lines[[1]]))
       lines <- c('%abc', lines)
    examples[[nm]] <- lines
}
```

We create a small helper function that creates symbols from all named items in 
a list. This is a little sketchy, since we do minimal error checking.  It also
does not work well embedded in other calls, since `parent.frame()` then returns 
the environment of the calling function, not the outer environment.
```{r}
make_vars <- function (.lst) {
  for (nm in names(.lst)) {
    if (! is.null(nm) && nchar(nm)>0) {
    assign(nm, .lst[[nm]], envir = parent.frame())
    }
  }
}
```

We create variables based on the names of the abc files.
```{r}
make_vars(examples)
```

Now, we add them all the the data folder as "*.rda" files.
The key step here is converting the names (which are strings from the for loop)
into symbols, which can be passed to `use_data()`.
```{r}
for (nm in names(examples)) {
  nm <- ensym(nm)
  do.call(usethis::use_data, list(nm, overwrite = TRUE))
}

```
